---
output: html_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

#### notes to self 

- check syntax of chsh at stan docs

<br><br><br><br><br><br>
<hr>

install routine below here
```{r eval=FALSE}
install.packages("rstan", repos="https://cloud.r-project.org/",
                 dependencies=TRUE)

```
```{r eval=FALSE}
fx <- inline::cxxfunction( signature(x = "integer", y = "numeric" ) , '
	return ScalarReal( INTEGER(x)[0] * REAL(y)[0] ) ;
' )
fx( 2L, 5 ) # should be 10
```
```{r}
library("rstan") # observe startup messages
```
As the startup message says, if you are using rstan locally on a multicore machine and have plenty of RAM to estimate your model in parallel, at this point execute
```{r eval=FALSE}
# may27/2017 didnt run these yet but revisit...
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

<hr><hr>
<br><br><br><br><br><br>

<hr><hr>

#### Example 1: Eight Schools

This is an example in Section 5.5 of Gelman et al (2003), which studied coaching effects from eight schools. For simplicity, we call this example "eight schools."

We start by writing a Stan program for the model and saving it in a new file 8schools.stan.

```{r}
# HERE IS eightschools.stan
```


In this model, we let theta be transformed parameters of mu and eta instead of directly declaring theta as parameters. By parameterizing this way, the sampler will run more efficiently (see detailed explanation). Assuming we have the 8schools.stan file in our working directory, we can prepare the data and fit the model as the following R code shows.

```{r}
schools_dat <- list(J = 8, 
                    y = c(28,  8, -3,  7, -1,  1, 18, 12),
                    sigma = c(15, 10, 16, 11,  9, 11, 10, 18))

fit <- stan(file = '8schools.stan', data = schools_dat, 
            iter = 1000, chains = 4)

```

We can also specify a Stan model using a character string by using argument model_code of function stan instead. However, this is not recommended.

The object fit, returned from function stan is an S4 object of class stanfit. Methods such as print, plot, and pairs are associated with the fitted result so we can use the following code to check out the results in fit. print provides a summary for the parameter of the model as well as the log-posterior with name lp__ (see the following example output). For more methods and details of class stanfit, see the help of class stanfit.

In particular, we can use extract function on stanfit objects to obtain the samples. extract extracts samples from the stanfit object as a list of arrays for parameters of interest, or just an array. In addition, S3 functions as.array and as.matrix are defined for stanfit object (using help("as.array.stanfit") to check out the help document in R).

```{r}
print(fit)
plot(fit)
pairs(fit, pars = c("mu", "tau", "lp__"))

la <- extract(fit, permuted = TRUE) # return a list of arrays 
mu <- la$mu 

### return an array of three dimensions: iterations, chains, parameters 
a <- extract(fit, permuted = FALSE) 

### use S3 functions as.array (or as.matrix) on stanfit objects
a2 <- as.array(fit)
m <- as.matrix(fit)

```

In addition, as in BUGS (or JAGS), CmdStan (the command line interface to Stan) needs all the data to be in an R dump file. In the case we have this file, rstan provides function read_rdump to read all the data into an R list. For example, if we have a file named "8schools.rdump" that contains the following text in our working directory.

```{r}
J <- 8
y <- c(28,  8, -3,  7, -1,  1, 18, 12)
sigma <- c(15, 10, 16, 11,  9, 11, 10, 18)

```

Then we can read the data from "8schools.rdump" as follows.

```{r}
schools_dat <- read_rdump('8schools.rdump')
```

The R dump file actually can be sourced using function source in R into the global environment. In this case, we can omit the data argument and stan will search the calling environment for objects that have the same names as in the data block of 8schools.stan. That is,

```{r}
source('8schools.rdump') 
fit <- stan(file = '8schools.stan', iter = 1000, chains = 4)
```


#### Example 2: Rats

The Rats example is also a popular example. For example, we can find the OpenBUGS version from here, which originally is from Gelfand et al (1990). The data are about the growth of 30 rats weekly for five weeks. In the following table, we list the data, in which we use x to denote the dates the data were collected. We can try this example using the linked data rats.txt and model code rats.stan.

```{r}
y <- as.matrix(read.table('https://raw.github.com/wiki/stan-dev/rstan/rats.txt', header = TRUE))
x <- c(8, 15, 22, 29, 36)
xbar <- mean(x)
N <- nrow(y)
T <- ncol(y)
rats_fit <- stan(file = 'https://raw.githubusercontent.com/stan-dev/example-models/master/bugs_examples/vol1/rats/rats.stan')
```

#### Example 3: Anything

You can run many of the BUGS examples and some others that we have created in Stan by executing

```{r}
model <- stan_demo()
```

and choosing an example model from the list that pops up. The first time you call stan_demo(), it will ask you if you want to download these examples. You should choose option 1 to put them in the directory where rstan was installed so that they can be used in the future without redownloading them. The model object above is an instance of class stanfit, so you can call print, plot, pairs, extract, etc. on it afterward.

<br><br><br><br><br><br>

##### VERBATIM PASTE OF THE RASMUS STAN CHEATSHEET 

###### A crash course to Stan’s syntax.

The basic syntax is similar to all “curly bracket” languages, such as C and JavaScript. But assignment and vectorization is similar to R.
```{stan output.var="blah", eval=FALSE}
// This is a comment (but # also works) 
# Here are some legal Stan statements
x = x + 1;
x = sqrt( (x + 1) * 3 );
int<upper=0> a[4];
if( x > 10) {
  x = 0;
}
# “;” is necessary after each statement.
```
As opposed to JavaScript, R and python, Stan is statically typed, and there are a lot of types specific to statistical modelling.
```{stan output.var="blah", eval=FALSE}
# Basic types
real x;
int y;
# Vectors are list of real numbers 
vector[10] v;      # A vector of length 10
# Matrices are tables of real numbers 
matrix[2,4] mat;   # A 2 by 4 matrix
# Arrays are lists of any data type
int a[4];          # A 4 length array of integers 
# Arrays can also be of higher dimension 
int a2[4, 3];      # A 4 by 3 array
```
All types can have constraints. Constraints are required for variables acting as parameters.
```{stan output.var="blah", eval=FALSE}
real mu; # No constraint implies [-∞, ∞]
real<lower=0> sigma; # [0, ∞]
int y<lower=0, upper=1>; # a “boolean”
# Works for vectors and arrays as well
vector<lower=1>[3] v;
int<upper=0> a[4];
# There are many “speciality” data types
simplex[3] p; # A vector of 3 positive
# reals that sums to 1.
corr_matrix[3] Sigma; # a 3 by 3
#correlation matrix
```
A Stan program consists of a number of blocks.
```{stan output.var="blah", eval=FALSE}
data { # the required data for the model
# Declarations ...
}
parameters { # the model’s parameters
# Declarations ...
}
model { # Defines the statistical model
# Declarations followed by statements ...
}
generated quantities {
# Declarations followed by statements ...
}
# But there are more block types...
```
Sampling statements define statistical relations
between parameters and data.
```{stan output.var="blah", eval=FALSE}
x ~ normal(mu, sigma); # Read as: x is
# distributed as a normal distribution
# with mean mu and SD sigma.
# There are many built in distribution.
mu ~ uniform(0, 100);
sigma ~ gamma(2, 2);
x1 ~ student_t(nu, mu, sigma);
lambda ~ exponential(1);
y ~ poisson(lambda);
p ~ beta(a, b);
s ~ binomial(30, p);
```
As in R, many functions are vectorized.
Here assuming v1 and v2 are vectors.
```{stan output.var="blah", eval=FALSE}
sum(v1); # Sums all elements in v;
v1 + v2; # The vector of the pairwise sum
# of v1 and v2
# Assume all elements of v are normal.
v ~ normal(mu, sigma);
# This is equivalent to using the for loop:
for (i in 1:n) {
v[i] ~ normal(mu, sigma);
}
# Note that “[]” is used to index vectors
# and arrays, and that 1 is the first index.
```
A minimal Stan program implementing a binomial
model.
```{stan output.var="blah", eval=FALSE}
data {
int n;
int x;
}
parameters {
real<lower=0, upper=1> p;
}
model {
p ~ uniform(0, 1);
x ~ binomial(n, p);
}
```
Running a Stan program is usually done from
another language such as Python or R. 
(Here assuming model_string contains the model
from the last slide.)
```{r eval=FALSE}
library(rstan)
data_list <- list(n = 30, x = 10)
s <- stan(model_code = model_string,
data = data_list)
```
```{python eval=FALSE}
import pystan
data_list = dict(n = 30, x = 10)
s = pystan.stan(model_code = model_string,
data = data_list)
```

<hr>

<br><br><br><br><br><br>

<hr>

Below here start writing code...

```{stan output.var="stn_code", eval=FALSE}
### FEED THE CODE BELOW TO model_code= ARGUMENT 
### IN CALL TO rstan::stan() FROM R 

// *x* a binomial var w *n* trials
data {
  int n;
  int x;
}

// *p* is proportion of sucesses for *x*
parameters {
  real<lower=0, upper=1>p;
}

// *p* uniform across [0,1]
// *x* binomial w *n* trials at *p* chance 
model {
  p ~ uniform(0, 1); 
  x ~ binomial(n, p);
}
```

```{r eval=FALSE}
library("rstan")
# the code blocks above, as a string
stn_code <- "<INSERT STAN CODE FROM CHUNK ABOVE>"
# "parameters" evaluated by `stan_code`
dat_list <- list(n=30, x=10)
# execute the model in stan_code, relative to `dat_list`
stan_mod <- stan(model_code=stn_code, data=dat_list)
```

